Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-08-03 analysis/paper/methods.Rmd
updated text

From the two-dimensional scans, we calculated pleiotropy likelihood ratio test statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

\begin{equation}
LOD = \log_{10} \left(\frac{\max_{\lambda_1, \lambda_2}L(B, \Sigma, \lambda_1, \lambda_2)}{\max_{\lambda}L(B, \Sigma, \lambda, \lambda)} \right)
\end{equation}{#eq:lod}

\begin{equation}
ll(\lambda_1, \lambda_2)
\end{equation}

@@

A mediation analysis in systems genetics begins with an identified QTL-expression trait 
association [@chick2016defining]. In our case, we choose a QTL that affects a nonlocal trait, then ask if any of 
the nearby local traits lie in the causal pathway from QTL to nonlocal trait.

involved fitting four linear models (Equations \@ref(eq:model1), \@ref(eq:model2), \@ref(eq:model3), \@ref(eq:model4)) [@chick2016defining].

@@

In all of our mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship. We only considered as potential mediators those local traits with strong univariate LODs.
In Equations~\ref{model1} through \ref{model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator, $\beta$ is the effect
of the mediator on the complex trait, 1 is a vector with all entries set to 1, $\alpha$
is the mean trait value, and $E$ is a vector of normally distributed random errors.

@@

### Heatmaps

>>>>

From the two-dimensional scans, we calculated pleiotropy likelihood ratio test (LOD) statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

\begin{equation}
LOD = \log_{10} \left(\frac{\max_{\lambda_1, \lambda_2}L(B, \Sigma, \lambda_1, \lambda_2)}{\max_{\lambda}L(B, \Sigma, \lambda, \lambda)} \right)
(\#eq:lod)
\end{equation}

@@

A mediation analysis in systems genetics begins with an identified QTL-expression trait 
association [@chick2016defining]. In our case, we choose a QTL that affects a nonlocal expression trait, then ask if any of 
the nearby local expression traits lie in the causal pathway from QTL to nonlocal trait.

A single mediation analysis involving one nonlocal trait and one local trait requires fitting four linear models (Equations \@ref(eq:model1), \@ref(eq:model2), \@ref(eq:model3), \@ref(eq:model4)) [@chick2016defining].

@@

In all of our mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship. 
We only considered as potential mediators those local traits with strong univariate LODs.
In Equations~\ref{model1} through \ref{model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator (a local expression trait, in our case), $\beta$ is the effect
of the mediator on the complex trait (after adjusting for other covariates in the model), 1 is a vector with all entries set to 1, $\alpha$
is a parameter that corresponds to the mean trait value (after adjusting for other covariates in the model), and $E$ is a vector of normally distributed random errors.

We included four covariates in our models: three wave indicators and sex. The wave indicators are binary vectors to account for wave membership. @keller2018genetic phenotyped the mice in four waves of nearly 100 mice per wave.

@@

### 2019-08-03 analysis/paper/results.Rmd
updated text

## Results

@@

Complex hotspots: Chr 5; Simple hotspots: Chr 2, Chr 13

Undetermined (right now): Chr 7, Chr 11

@@

Examinating the ranges of pleiotropy test statistics within a hotspot also informs our conclusions. 
Some hotspots have maximum pleiotropy test statistics under 6 (Figure \@ref(fig:scatter-7)), while 
other hotspots have local-nonlocal pairs with pleiotropy test statistics over 15 (Figure 
\@ref(fig:scatter-2)).

>>>>

# Results

@@

Complex hotspots: Chr 5; Simple hotspots: Chr 2, Chr 13, Chr 7

Undetermined (right now): Chr 11

@@

Examining the ranges of pleiotropy test statistics within a hotspot also informs our conclusions. 
Some hotspots have maximum pleiotropy test statistics under 6 (Figure \@ref(fig:scatter-7)), while 
other hotspots have local-nonlocal pairs with pleiotropy test statistics over 15 (Figure 
\@ref(fig:scatter-2)).
